Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-12 article.Rmd
Placeholder images

# TRUE to show code and warnings; FALSE to remove them for publication
display_code_blocks(TRUE)

@@

Many other American names, such as Madison, Morgan, Sydney, and Kendall, have also changed gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely moves in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name "Madison" changed from male to female suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity, and by the early 2000s it was one of the ten most frequent names for baby girls. Broadly speaking, since the late twentieth century Americans girls are more likely than American boys to receive a "trendy" name, and once that name is associated with females it is unlikely to be given to male babies. The few exceptions of names which go from being female to male (such as "Jan") have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change. **this last bit about "Jan" might just muddle things - unless we have a clear explanation or hypothesis as to what does explain the change, not sure it's worth highlighting**

@@

[chart – raw numbers]

@@

[chart - proportion]

@@

```{r gender-book-authors}
ahr <- read.csv("data/ahr.utf8.csv", stringsAsFactors = FALSE) %>%
  tbl_df()

@@

But a closer look shows important differences between the proportion of women writing history dissertations and the proportion of women appearing in the AHR. Although both have trended upwards since the 1970s the slope is a lot steeper for dissertations than it is for the AHR. Female historians have very nearly closed the gaps in terms of newly completed dissertations, but the glass ceiling remains stubbornly low in terms of what happens from that point onwards. In book reviews published in the AHR male authors continue to outnumber female authors by a factor of nearly 2 to 1. The long-term gains made by female doctoral students do not carry over in the pages of the historical profession’s leading journal.

[chart comparing dissertations with AHR]

@@

[chart: male authored books]
[chart: female authored books]

@@

[chart – sub-fields]

>>>>

# TRUE to show code and warnings; FALSE to remove them for publication
display_code_blocks(FALSE)

@@

Many other American names, such as Madison, Morgan, Sydney, and Kendall, have also changed gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely moves in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name "Madison" changed from male to female suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity, and by the early 2000s it was one of the ten most frequent names for baby girls. Broadly speaking, since the late twentieth century Americans girls are more likely than American boys to receive a "trendy" name, and once that name is associated with females it is unlikely to be given to male babies.^[The few exceptions of names which go from being female to male (such as "Jan") have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

![Number of dissertations by gender](img/plot-gender.png)

@@

![Proportion of dissertations for each gender](img/dissertations-by-gender-proportion.png)

@@

![Gender of authors of books reviewed in the AHR](img/book-authors.jpg)

```{r gender-book-authors, eval=FALSE}
ahr <- read.csv("data/ahr.utf8.csv", stringsAsFactors = FALSE) %>%
  tbl_df()

@@

But a closer look shows important differences between the proportion of women writing history dissertations and the proportion of women appearing in the AHR. Although both have trended upwards since the 1970s the slope is a lot steeper for dissertations than it is for the AHR. Female historians have very nearly closed the gaps in terms of newly completed dissertations, but the glass ceiling remains stubbornly low in terms of what happens from that point onwards. In book reviews published in the AHR male authors continue to outnumber female authors by a factor of nearly 2 to 1. The long-term gains made by female doctoral students do not carry over in the pages of the historical profession's leading journal.

![Gap between female authorship of dissertations and books](img/book-dissertation-gap.jpg)

@@

![Gender of reviewers of books authored by women](img/female_authors_withreviewers.jpg)

@@

![Gender of reviewers by category of review](img/categories_gender_bytime.jpg)
